---
title: "Getting started with egrandis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with egrandis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

```{r}
library(egrandis)
```

`egrandis` reproduces the INIA SAG grandis 2021 system for projecting
*Eucalyptus grandis* stands in subtropical South America. This vignette
walks through the public API.

## A simple unthinned projection

The reference scenario from INIA Uruguay: Zone 7 (Tacuarembó / Rivera),
site index 30 m at base age 10, 550 trees per hectare, projected from
age 1 to age 16.

```{r}
sim <- simulate_inia(
  SI    = 30,
  N0    = 550,
  G0    = 1.7,
  Hd0   = 5.2,
  dmax0 = 8.0,
  SDd0  = 1.3,
  t0    = 1,
  t_end = 16,
  zone  = 7
)

head(sim$trajectory)
```

A formatted summary including the trajectory table is available through
`inia_print_summary()`:

```{r, results = "hide"}
inia_print_summary(sim)
```

## Adding thinning events

`thins` accepts a list of events, each a list of `age` (year of the
thin) and `N_after` (post-thin trees per hectare). The simulator applies
the empirical SAG 2003 BA-reduction equation to compute residual basal
area and tracks volume removed.

```{r}
sim_thinned <- simulate_inia(
  SI    = 30, N0 = 550, G0 = 1.7,
  Hd0   = 5.2, dmax0 = 8.0, SDd0 = 1.3,
  t0    = 1, t_end = 12, zone = 7,
  thins = list(
    list(age = 3, N_after = 412),
    list(age = 7, N_after = 197)
  )
)

sim_thinned$thinnings
sim_thinned$total_yield  # standing + thinned volume
```

## Comparing density regimes

Because the INIA model is a whole-stand projection, comparing regimes is
just a matter of calling `simulate_inia()` with different inputs. Here
we sweep across four initial densities and plot MAI curves.

```{r}
densities <- c(400, 550, 800, 1111)

# Match INIA's age-1 BA convention: solve from a starting Dq of 6.3 cm.
g_from_dq <- function(N, Dq = 6.3) N * pi * (Dq / 200)^2

sims <- lapply(densities, function(n) {
  simulate_inia(
    SI = 30, N0 = n, G0 = g_from_dq(n),
    Hd0 = 5.2, dmax0 = 8.0, SDd0 = 1.3,
    t0 = 1, t_end = 16, zone = 7
  )
})

plot(NULL, xlim = c(1, 16), ylim = c(0, 55),
     xlab = "Age (years)", ylab = "MAI (m^3/ha/yr)",
     main = "MAI by initial density (Zone 7, SI = 30)")
cols <- c("steelblue", "darkgreen", "darkorange", "firebrick")
for (i in seq_along(sims)) {
  lines(sims[[i]]$trajectory$age, sims[[i]]$trajectory$MAI,
        col = cols[i], lwd = 2)
}
legend("bottomright", legend = paste(densities, "TPH"),
       col = cols, lwd = 2, bty = "n")
```

## Diameter distributions

The simulator carries enough stand-level moments (quadratic mean
diameter, maximum diameter, diameter SD) to recover an inverse Weibull
diameter distribution at any age. `inia_get_distribution()` is the
convenience wrapper:

```{r}
dd10 <- inia_get_distribution(sim, age = 10)
head(dd10)

plot(dd10$class_mid, dd10$freq, type = "h", lwd = 4,
     xlab = "DBH (cm)", ylab = "Trees / ha",
     main = "Recovered diameter distribution at age 10")
```

The underlying recovery function `inia_diam_dist()` can also be called
directly given any set of stand moments.

## The validation dataset

`sag_validation` is the dataset of five reference trajectories captured
from the SAG grandis 2021 online simulator. Use it to verify the
package or as a benchmark when re-fitting parameters.

```{r}
data(sag_validation)
names(sag_validation)

# Reproduce the reference scenario and compare basal-area trajectories
ref <- sag_validation$z7_si30_n550
sim <- do.call(
  simulate_inia,
  modifyList(ref$inputs, list(t0 = 1, t_end = 16,
                              DAP_medio0 = NULL))
)

cbind(
  age      = ref$trajectory$age,
  sag_AB   = ref$trajectory$AB,
  sim_AB   = sim$trajectory$AB,
  delta    = round(sim$trajectory$AB - ref$trajectory$AB, 2)
)
```

The `tests/` directory builds on this dataset to encode the documented
per-submodel tolerances.

## Taper, tree-level volume, and merchantable assortment

`egrandis` ships with a tree-level taper-volume system separate from the
INIA stand-level volume. It uses the Fang et al. (2000) compatible
segmented system with the *E. grandis* coefficients reported by Hirigoyen
et al. (2021) (iForest 14: 127--136, Table 4), fitted to felled-tree taper
data from Uruguay.

```{r}
# Diameter profile of an average mid-rotation tree
D <- 25; H <- 28
h <- seq(0.10, H, length.out = 200)
d <- inia_taper(h, D = D, H = H)

plot(h, d, type = "l", lwd = 2,
     xlab = "Height above ground (m)", ylab = "Stem diameter (cm)",
     main = sprintf("Fang taper for D=%dcm, H=%dm", D, H))
abline(h = c(8, 14, 25), lty = 3, col = "grey50")
text(rep(0.5, 3), c(8, 14, 25), c("8 cm", "14 cm", "25 cm"),
     pos = 3, col = "grey30", cex = 0.8)
```

Single-tree volumes:

```{r}
inia_tree_total_vol(D = 25, H = 28)         # closed-form Fang total
inia_tree_vol(D = 25, H = 28, d_top = 8)    # merch to an 8 cm top
inia_tree_vol(D = 25, H = 28, d_top = 14)   # merch to a 14 cm top
inia_height_at_d(d_top = 14, D = 25, H = 28) # height where diameter = 14 cm
```

The cone-tail closure of segment 3 means integrated total volume is
within ~1% of the closed-form Fang total; for total volume always
prefer `inia_tree_total_vol()`.

### Stand-level merchantable volume by product

`inia_merch_vol()` runs the taper across the recovered Weibull diameter
distribution at the requested age and buckets the volume into named
products. By default it splits into veneer (>= 25 cm), solid (>= 14 cm),
and pulp (>= 8 cm) on a butt-up cascade — the largest small-end limit
takes the butt log down to its limit, the next product takes the next
section, and so on.

```{r}
mv <- inia_merch_vol(sim, age = 16)
attr(mv, "totals")

barplot(
  c(veneer = unname(attr(mv, "totals")["vol_veneer"]),
    solid  = unname(attr(mv, "totals")["vol_solid"]),
    pulp   = unname(attr(mv, "totals")["vol_pulp"])),
  col = c("steelblue", "darkorange", "olivedrab"),
  ylab = "Volume (m^3/ha)",
  main = "Product assortment at age 16 (Zone 7, SI=30, 550 TPH)"
)
```

Pass a custom `products =` list for any assortment scheme:

```{r}
mv_custom <- inia_merch_vol(
  sim, age = 16,
  products = list(saw = list(d_min = 18), pulp = list(d_min = 8))
)
attr(mv_custom, "totals")
```

### Two volume estimators, two purposes

`egrandis` has *two* volume sources that do not agree exactly:

- `simulate_inia()` reports stand-level total volume (`Vol_Total`) from
  the INIA equation `V = exp(b0 + b1*ln(G*Hd) + b2*(Hd/Dg) + b3*Z7)`,
  refit to SAG 2021 online output. This is the right number to compare
  against the SAG simulator.
- `inia_tree_total_vol()` and `inia_merch_vol()` use the Fang taper-volume
  system fit by Hirigoyen et al. to felled-tree measurements. This is
  the right number for product assortments and log bucking.

The two estimates can differ by 10-20% at the stand level because they
reflect different data sources and bark conventions. Use the one that
matches your downstream calculation.

## Known limitations

The current INIA implementation has several documented limitations that
users should keep in mind. See `?simulate_inia` and the README for
detail:

- Basal area overshoots when initialized at age 1 well below the
  Schumacher asymptote. Initialize from age 3+ with measured G for
  reliable projections.
- Zone 8/9 mortality is unfitted (the current parameters produce no
  background mortality). Zone 7 is the validated case.
- The dmax submodel was fit at Zone 7 / 550 TPH. High-density and
  Zone 8/9 dmax trajectories drift from SAG.
- Mortality has no density feedback. Self-thinning at high density is
  not represented.
- The Fang taper's segment 3 (above relative height ~0.94) is replaced
  with a linear cone tail to keep diameter monotone. The displaced
  volume is < 1% of total tree volume but means integrated total volume
  is a hair below the closed-form Fang total.
