---
title: "Getting started with egrandis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with egrandis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

```{r}
library(egrandis)
```

`egrandis` implements the INIA SAG grandis 2021 stand-projection system
for *Eucalyptus grandis*, together with a Fang (2000) compatible taper /
merchantable-volume system and an aboveground biomass / carbon module.
This vignette walks through the public API.

## A simple unthinned projection

A whole-stand projection takes site index, initial stocking, and the
stand state at the start age. Here we start from a healthy young stand
at age 2 in Zone 7.

```{r}
sim <- simulate_inia(
  SI    = 28,
  N0    = 900,
  G0    = 7.0,
  Hd0   = 7.0,
  dmax0 = 13.0,
  SDd0  = 1.8,
  t0    = 2,
  t_end = 16,
  zone  = 7
)

head(sim$trajectory)
```

A formatted summary including the trajectory table is available through
`inia_print_summary()`:

```{r, results = "hide"}
inia_print_summary(sim)
```

## Adding thinning events

`thins` accepts a list of events, each a list of `age` (year of the
thin) and `N_after` (post-thin trees per hectare). The simulator applies
the empirical SAG 2003 BA-reduction equation to compute residual basal
area and tracks volume removed.

```{r}
sim_thinned <- simulate_inia(
  SI    = 28, N0 = 900, G0 = 7.0,
  Hd0   = 7.0, dmax0 = 13.0, SDd0 = 1.8,
  t0    = 2, t_end = 14, zone = 7,
  thins = list(
    list(age = 4, N_after = 600),
    list(age = 9, N_after = 300)
  )
)

sim_thinned$thinnings
sim_thinned$total_yield  # standing + thinned volume
```

## Comparing density regimes

Because the INIA model is a whole-stand projection, comparing regimes is
just a matter of calling `simulate_inia()` with different inputs. Here
we sweep across four initial densities and plot MAI curves.

```{r}
densities <- c(800, 900, 1000, 1100)

# Initial BA implied by a starting Dq of ~10 cm at age 2.
g_from_dq <- function(N, Dq = 10) N * pi * (Dq / 200)^2

sims <- lapply(densities, function(n) {
  simulate_inia(
    SI = 28, N0 = n, G0 = g_from_dq(n),
    Hd0 = 7.0, dmax0 = 13.0, SDd0 = 1.8,
    t0 = 2, t_end = 16, zone = 7
  )
})

plot(NULL, xlim = c(2, 16), ylim = c(0, 50),
     xlab = "Age (years)", ylab = "MAI (m^3/ha/yr)",
     main = "MAI by initial density (Zone 7, SI = 28)")
cols <- c("steelblue", "darkgreen", "darkorange", "firebrick")
for (i in seq_along(sims)) {
  lines(sims[[i]]$trajectory$age, sims[[i]]$trajectory$MAI,
        col = cols[i], lwd = 2)
}
legend("bottomright", legend = paste(densities, "TPH"),
       col = cols, lwd = 2, bty = "n")
```

## Diameter distributions

The simulator carries enough stand-level moments (quadratic mean
diameter, maximum diameter, diameter SD) to recover an inverse Weibull
diameter distribution at any age. `inia_get_distribution()` is the
convenience wrapper:

```{r}
dd10 <- inia_get_distribution(sim, age = 10)
head(dd10)

plot(dd10$class_mid, dd10$freq, type = "h", lwd = 4,
     xlab = "DBH (cm)", ylab = "Trees / ha",
     main = "Recovered diameter distribution at age 10")
```

The underlying recovery function `inia_diam_dist()` can also be called
directly given any set of stand moments.

## The validation dataset

`sag_validation` is the dataset of five reference trajectories captured
from the SAG grandis 2021 online simulator. Use it to verify the
package or as a benchmark when re-fitting parameters.

```{r}
data(sag_validation)
names(sag_validation)

# Reproduce one of the reference scenarios and compare basal-area
# trajectories on the same inputs SAG 2021 used.
ref <- sag_validation$z7_si30_n550
sim_ref <- do.call(
  simulate_inia,
  modifyList(ref$inputs, list(t0 = 1, t_end = 16,
                              DAP_medio0 = NULL))
)

cbind(
  age      = ref$trajectory$age,
  sag_AB   = ref$trajectory$AB,
  sim_AB   = sim_ref$trajectory$AB,
  delta    = round(sim_ref$trajectory$AB - ref$trajectory$AB, 2)
)
```

The `tests/` directory builds on this dataset to encode the documented
per-submodel tolerances.

## Taper, tree-level volume, and merchantable assortment

`egrandis` ships with a tree-level taper-volume system separate from the
INIA stand-level volume. It uses the Fang et al. (2000) compatible
segmented system with the *E. grandis* coefficients reported by Hirigoyen
et al. (2021) (iForest 14: 127--136, Table 4), fitted to felled-tree taper
data from Uruguay.

```{r}
# Diameter profile of an average mid-rotation tree
D <- 25; H <- 28
h <- seq(0.10, H, length.out = 200)
d <- inia_taper(h, D = D, H = H)

plot(h, d, type = "l", lwd = 2,
     xlab = "Height above ground (m)", ylab = "Stem diameter (cm)",
     main = sprintf("Fang taper for D=%dcm, H=%dm", D, H))
abline(h = c(8, 14, 25), lty = 3, col = "grey50")
text(rep(0.5, 3), c(8, 14, 25), c("8 cm", "14 cm", "25 cm"),
     pos = 3, col = "grey30", cex = 0.8)
```

Single-tree volumes:

```{r}
inia_tree_total_vol(D = 25, H = 28)         # closed-form Fang total
inia_tree_vol(D = 25, H = 28, d_top = 8)    # merch to an 8 cm top
inia_tree_vol(D = 25, H = 28, d_top = 14)   # merch to a 14 cm top
inia_height_at_d(d_top = 14, D = 25, H = 28) # height where diameter = 14 cm
```

The cone-tail closure of segment 3 means integrated total volume is
within ~1% of the closed-form Fang total; for total volume always
prefer `inia_tree_total_vol()`.

### Stand-level merchantable volume by product

`inia_merch_vol()` runs the taper across the recovered Weibull diameter
distribution at the requested age and buckets the volume into named
products. Default assortment uses three generic solid-wood grades --
`large_sawlog` (>= 25 cm), `small_sawlog` (>= 14 cm), and `pulp`
(>= 8 cm) -- bucketed butt-up: the largest small-end limit takes the
butt log down to its limit, the next product takes the next section,
and so on.

```{r}
mv <- inia_merch_vol(sim, age = 16)
attr(mv, "totals")

barplot(
  c(`large_sawlog` = unname(attr(mv, "totals")["vol_large_sawlog"]),
    `small_sawlog` = unname(attr(mv, "totals")["vol_small_sawlog"]),
    `pulp`         = unname(attr(mv, "totals")["vol_pulp"])),
  col = c("steelblue", "darkorange", "olivedrab"),
  ylab = "Volume (m^3/ha)",
  main = "Product assortment at age 16"
)
```

Pass a custom `products =` list for any assortment scheme:

```{r}
mv_custom <- inia_merch_vol(
  sim, age = 16,
  products = list(sawlog = list(d_min = 18), pulp = list(d_min = 8))
)
attr(mv_custom, "totals")
```

### Two volume estimators, two purposes

`egrandis` has *two* volume sources that do not agree exactly:

- `simulate_inia()` reports stand-level total volume (`Vol_Total`) from
  the INIA equation `V = exp(b0 + b1*ln(G*Hd) + b2*(Hd/Dg) + b3*Z7)`,
  refit to SAG 2021 online output. This is the right number to compare
  against the SAG simulator.
- `inia_tree_total_vol()` and `inia_merch_vol()` use the Fang taper-volume
  system fit by Hirigoyen et al. to felled-tree measurements. This is
  the right number for product assortments and log bucking.

The two estimates can differ by 10-20% at the stand level because they
reflect different data sources and bark conventions. Use the one that
matches your downstream calculation.

## Aboveground biomass, carbon, and CO2-equivalent

`egrandis` also estimates total aboveground dry biomass (stem +
branches + foliage) using an allometric equation
\(AGB = a \cdot d^{b_1} \cdot h^{b_2}\) integrated across the recovered
diameter distribution, with a stand-specific exponential h-d curve. The
coefficients were jointly fitted against three SAG 2021 reference
scenarios (RRMSE 5.4%). Carbon and CO2-equivalent follow IPCC
subtropical-hardwood defaults.

```{r}
sim <- inia_add_biomass(sim)
head(sim$trajectory[, c("age", "Vol_Total", "Biomasa", "Carbon", "CO2eq")])
```

```{r}
op <- par(mar = c(4, 4.2, 1, 4.2))
with(sim$trajectory, {
  plot(age, Biomasa, type = "l", lwd = 2, col = "darkgreen",
       xlab = "Age (years)", ylab = "Biomass (t/ha)",
       main = "Aboveground biomass and CO2 stock")
  par(new = TRUE)
  plot(age, CO2eq, type = "l", lwd = 2, col = "steelblue", lty = 2,
       axes = FALSE, xlab = "", ylab = "")
  axis(4, col.axis = "steelblue", col = "steelblue")
  mtext("CO2 equivalent (t CO2/ha)", side = 4, line = 2.8, col = "steelblue")
  legend("topleft", legend = c("AGB (t/ha)", "CO2eq (t CO2/ha)"),
         col = c("darkgreen", "steelblue"), lwd = 2, lty = c(1, 2), bty = "n")
})
par(op)
```

For one-off estimates without a full simulation, call `inia_stand_agb()`
directly with stand-level moments:

```{r}
inia_stand_agb(N = 700, Dq = 28.0, dmax = 35.0, SDd = 4.5, Hd = 24.0)
```

## Known limitations

The current INIA implementation has several documented limitations that
users should keep in mind. See `?simulate_inia` and the README for
detail:

- Basal area overshoots when initialized at age 1 well below the
  Schumacher asymptote. Initialize from age 2 or later with a measured
  G0 for reliable projections.
- Zone 8/9 mortality is unfitted (the current parameters produce no
  background mortality). Zone 7 is the validated case.
- The dmax submodel was fit at one density / zone combination. High
  densities and Zone 8/9 dmax trajectories drift from SAG.
- Mortality has no density feedback. Self-thinning at high density is
  not represented.
- The Fang taper's segment 3 (above relative height ~0.94) is replaced
  with a linear cone tail to keep diameter monotone. The displaced
  volume is < 1% of total tree volume but means integrated total volume
  is a hair below the closed-form Fang total.
- The biomass module's worst calibration error is around -15% on the
  youngest / lowest-site / lowest-density case. Ages 7+ inside the
  calibration envelope are within +-8%.
